---
title: |
  | Estimated Annual Agricultural Pesticide Use
  | Pacific Northwest (WA/OR/ID), Regional Ranking 2014-2018
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  bookdown::pdf_document2:
    number_sections: no
    toc: no
geometry: margin = 0.75in
fontsize: 10pt
link-citations: true
linkcolor: blue
urlcolor: blue
---

```{r knitr.global.options, include=F}
knitr::opts_chunk$set(echo=F,
                      warning=F,
                      fig.align='center',
                      fig.pos='H',
                      fig.width=9,
                      fig.height=6,
                      out.width='100%',
                      fig.path='Figs/',
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)

options(scipen=999)
```

```{r load.packages, include=F}
required_pkgs <- c("knitr", "bookdown", "dplyr", "tidyr", "readr", "ggplot2", "scales", "kableExtra")
invisible(lapply(required_pkgs, function(pkg) library(pkg, character.only = TRUE)))
```

```{r constants}
region_fips  <- c("53", "41", "16")
region_names <- c("53" = "Washington", "41" = "Oregon", "16" = "Idaho")
region_abbrevs <- c("53" = "WA", "41" = "OR", "16" = "ID")
region_label <- "Pacific Northwest (WA/OR/ID)"

# 2019 is excluded: the county-level file available at the time of writing
# is a partial/preliminary release covering only 69 compounds nationwide
# (vs. ~400-410 in every other year, 1992-2018), for all three states
# alike, so anchoring the "most recent year" on it would silently drop
# hundreds of real compounds from every ranking table. Re-run with
# 2015-2019 once a complete 2019 release is available.
window_years <- 2014:2018
recent_year  <- max(window_years)

cat_hex <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
              "#e87ba4", "#008300", "#4a3aa7", "#e34948")
muted_grey <- "#898781"
type_levels <- c("Insecticide", "Herbicide", "Fungicide", "Fumigant",
                  "Plant Growth Regulator", "Biological", "Other")
type_colors <- setNames(c(cat_hex[1:6], muted_grey), type_levels)
```

```{r upload.and.clean.data, include=F}

## Read 2014-2018 county-level estimates from ../estimates/EPest.county.estimates.<year>.txt,
## keeping only WA/OR/ID counties.

read_year <- function(yr) {
  read_delim(
    sprintf("../estimates/EPest.county.estimates.%d.txt", yr),
    delim = "\t",
    col_types = cols(
      COMPOUND = col_character(),
      YEAR = col_integer(),
      STATE_FIPS_CODE = col_character(),
      COUNTY_FIPS_CODE = col_character(),
      EPEST_LOW_KG = col_double(),
      EPEST_HIGH_KG = col_double()
    )
  ) %>%
    filter(STATE_FIPS_CODE %in% region_fips)
}

region_raw <- bind_rows(lapply(window_years, read_year))

classification <- read_csv("../data/pesticide_classification.csv", col_types = cols())

# Regional totals: sum across every WA + OR + ID county, per compound-year
region_yearly <- region_raw %>%
  group_by(COMPOUND, YEAR) %>%
  summarize(EPEST_LOW_KG = sum(EPEST_LOW_KG, na.rm = TRUE),
            EPEST_HIGH_KG = sum(EPEST_HIGH_KG, na.rm = TRUE), .groups = "drop")

# Each state's own total, same years, for the regional-vs-state-rank comparison
state_yearly <- lapply(region_fips, function(fips) {
  region_raw %>%
    filter(STATE_FIPS_CODE == fips) %>%
    group_by(COMPOUND, YEAR) %>%
    summarize(EPEST_LOW_KG = sum(EPEST_LOW_KG, na.rm = TRUE),
              EPEST_HIGH_KG = sum(EPEST_HIGH_KG, na.rm = TRUE), .groups = "drop")
})
names(state_yearly) <- region_fips

# Each window (1/3/5-yr) is ranked independently so a compound's presence
# in the appendix tables never depends on whether it also happens to
# appear in the *other* windows (e.g. a compound with real 2014-2018 use
# but zero recorded use in the single most recent year must still show up
# in the 5-yr appendix ranking).
compute_windows <- function(yearly) {
  one <- yearly %>%
    filter(YEAR == recent_year) %>%
    transmute(COMPOUND, EPEST_LOW_KG_1YR = EPEST_LOW_KG, EPEST_HIGH_KG_1YR = EPEST_HIGH_KG) %>%
    arrange(desc(EPEST_HIGH_KG_1YR)) %>%
    mutate(RANK_1YR = row_number())

  three <- yearly %>%
    filter(YEAR %in% tail(window_years, 3)) %>%
    group_by(COMPOUND) %>%
    summarize(EPEST_LOW_KG_3YR_AVG = mean(EPEST_LOW_KG), EPEST_HIGH_KG_3YR_AVG = mean(EPEST_HIGH_KG), .groups = "drop") %>%
    arrange(desc(EPEST_HIGH_KG_3YR_AVG)) %>%
    mutate(RANK_3YR = row_number())

  five <- yearly %>%
    group_by(COMPOUND) %>%
    summarize(EPEST_LOW_KG_5YR_AVG = mean(EPEST_LOW_KG), EPEST_HIGH_KG_5YR_AVG = mean(EPEST_HIGH_KG), .groups = "drop") %>%
    arrange(desc(EPEST_HIGH_KG_5YR_AVG)) %>%
    mutate(RANK_5YR = row_number())

  list(one = one, three = three, five = five)
}

region_w <- compute_windows(region_yearly)
state_w  <- lapply(state_yearly, compute_windows)

top_n <- 80

# The main summary table (Table 2) intentionally requires a compound to
# appear in all three regional windows at once, since it displays 1/3/5-yr
# figures side by side for the same compound; the appendix tables below do
# not use this joined version.
region_ranked <- region_w$one %>%
  inner_join(region_w$three, by = "COMPOUND") %>%
  inner_join(region_w$five, by = "COMPOUND") %>%
  mutate(across(where(is.numeric) & !starts_with("RANK"), ~round(.x)))

state_rank_cols <- Reduce(function(acc, fips) {
  col <- paste0("RANK_1YR_", region_abbrevs[fips])
  acc %>% left_join(
    state_w[[fips]]$one %>% select(COMPOUND, !!col := RANK_1YR),
    by = "COMPOUND"
  )
}, region_fips, region_ranked %>% arrange(RANK_1YR) %>% slice(1:top_n))

region_top <- state_rank_cols %>%
  left_join(classification, by = "COMPOUND") %>%
  mutate(CLASS = coalesce(CLASS, "Other/Unclassified"),
         ABBREV = coalesce(ABBREV, "OTH"),
         TYPE = coalesce(TYPE, "Other"),
         TYPE = factor(TYPE, levels = type_levels)) %>%
  select(RANK_1YR, COMPOUND, CLASS, ABBREV, TYPE, EPEST_HIGH_KG_1YR, EPEST_HIGH_KG_3YR_AVG, EPEST_HIGH_KG_5YR_AVG,
         starts_with("RANK_1YR_"))
```

# Summary

Table \@ref(tab:merged-tables) lists the rank and best available estimates of
mass applied (kg) for the top `r top_n` agricultural pesticides across
**`r region_label`** during `r recent_year`, as reported by [USGS Estimated
Annual Agricultural Pesticide
Use](https://water.usgs.gov/nawqa/pnsp/usage/maps/county-level/) data,
summed across every county in Washington, Oregon, and Idaho. The table
also includes rank by 3-year average
(`r paste(range(tail(window_years,3)), collapse="-")`) and 5-year average
(`r paste(range(window_years), collapse="-")`) estimates of mass applied,
each compound's chemical class (abbreviated; see the key in Table
\@ref(tab:class-key) and the full classification in
`data/pesticide_classification.csv`, shared with each state's own
chapters), and — for comparison — its `r recent_year` rank *within each
state considered alone*. Rankings and estimates are based on the
*EPest-high method* described by [Thelin and Stone
(2013)](https://pubs.usgs.gov/sir/2013/5009/) and [Baker and Stone
(2015)](https://pubs.usgs.gov/ds/0907/):

> "EPest-low and EPest-high, are used to estimate a range of pesticide use.
> Both EPest-low and EPest-high methods incorporate proprietary surveyed
> rates for Crop Reporting Districts (CRDs), but EPest-low and EPest-high
> estimates differ in how they treat situations when a CRD was surveyed and
> pesticide use was not reported for a particular crop present in the CRD.
> In these situations, EPest-low assumes zero use in the CRD for that
> pesticide-by-crop combination. EPest-high, however, treats the unreported
> use for that pesticide-by-crop combination in the CRD as missing data. In
> this case, pesticide-by-crop use rates from neighboring CRDs or CRDs
> within the same region are used to estimate the pesticide-by-crop
> EPest-high rate for the CRD." [*-Pesticide National Synthesis Project
> webpage*](https://water.usgs.gov/nawqa/pnsp/usage/maps/about.php)

USGS lists several caveats of these data, in particular:

* Reliability decreases with scale (e.g. "detailed interpretation of where
  and how much use occurs within a county is not appropriate"); summing
  three states together inherits this caveat at an even larger scale.
* EPest-low estimates are more likely to reflect state-based restrictions
  on pesticide use than EPest-high estimates.
* A blank cell in the source file (no reported/estimated use for that
  compound-year-county) is treated as zero in this chapter.
* **2019 is excluded from this ranking.** The 2019 county-level file
  available at the time of writing covers only 69 compounds nationwide,
  versus roughly 400 compounds in every other year from 1992-2018 — a
  partial/preliminary release, not a real drop in the number of pesticides
  in use. Anchoring the ranking on it would silently omit hundreds of real
  compounds. This chapter uses `r recent_year` (the most recent complete
  year) as the 1-year snapshot instead; re-run with `r recent_year+1`
  included once a complete 2019 release is available.
* **Comparing raw mass across states is not apples to apples** — WA, OR,
  and ID farm very different total acreages and crop mixes, so a
  compound's regional rank reflects total farmland as much as anything
  else. The per-state rank columns below are about *where a compound
  concentrates*, not a claim about which state uses more pesticide overall.

A compound's regional rank and its individual state ranks can differ
substantially — a compound used heavily across Idaho's potato acreage but
barely at all in Oregon will rank high regionally only if Idaho's volume
carries it; `NA` in a state's rank column means the compound was not among
that state's own top `r top_n` for `r recent_year`.

```{r class-key, include=T}
# ABBREV -> CLASS is a clean 1:1 mapping (each of the 30 classes has one
# code); ABBREV -> TYPE is not (e.g. Organophosphate spans both Insecticide
# and Herbicide, since bensulide and tribufos are OP herbicides), so the key
# only shows Code/Class -- adding Type would print each code multiple times.
kable(classification %>% distinct(ABBREV, CLASS) %>% arrange(CLASS),
      "latex", booktabs = T, longtable = TRUE,
      col.names = c("Code", "Chemical class"),
      caption = "Chemical class abbreviation key (used in the Class column of Table 2)") %>%
  kable_styling(latex_options = c("striped", "repeat_header", "HOLD_position"), font_size = 8)
```

```{r merged-tables, include=T}
# NOTE: the 80-row table does not fit on one page. kableExtra's usual
# "scale_down" (latex \resizebox) bundles the whole table into one atomic
# box that cannot break across pages, which silently drops every row past
# the page boundary instead of erroring. Use longtable (page-breaking)
# with a smaller font and abbreviated class codes instead.
display <- region_top %>% select(-CLASS, -TYPE)
names(display) <- c("Rank (PNW)", "Compound", "Class", "EPest-high 1yr",
                     "EPest-high 3yr avg", "EPest-high 5yr avg",
                     paste0("Rank (", region_abbrevs[region_fips], ")"))

# 9 columns (3 more than the single-state chapters, for the 3 per-state
# rank columns) don't fit in portrait even at font_size 7 -- the two
# rightmost columns ran off the page edge. Landscape gives ~9.5in of
# usable width instead of ~7in, which comfortably fits all 9 at font_size 8.
kable(display, "latex", booktabs = T, digits = 0, longtable = TRUE,
      caption = paste0("Top ", top_n, " EPest-high Estimates, ", region_label, ", 1-Yr (", recent_year,
                        "), 3-Yr Avg, 5-Yr Avg, with Chemical Class (abbreviated; see the ",
                        "classification key table above) and Each State's Own Rank")) %>%
  kable_styling(latex_options = c("striped", "repeat_header", "HOLD_position"), font_size = 8) %>%
  landscape()

write.csv(region_top, "pnw.regional.pesticide.use.2014-2018.csv", row.names = FALSE)
```

```{r top-chart, include=T, fig.cap="Top 25 compounds region-wide by EPest-high mass applied, most recent year, colored by pesticide type."}
top_chart <- region_top %>%
  slice(1:25) %>%
  mutate(COMPOUND = forcats::fct_reorder(COMPOUND, EPEST_HIGH_KG_1YR))

ggplot(top_chart, aes(x = COMPOUND, y = EPEST_HIGH_KG_1YR, fill = TYPE)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = type_colors, name = "Type", drop = FALSE) +
  scale_y_continuous(labels = comma) +
  labs(title = paste0(region_label, " top 25 pesticides, ", recent_year),
       x = NULL, y = "EPest-high estimate (kg)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

# Appendices

```{r table-1yr, include=T}
kable(head(region_w$one %>% arrange(RANK_1YR) %>%
             select(RANK_1YR, COMPOUND, EPEST_LOW_KG_1YR, EPEST_HIGH_KG_1YR), n = 50),
      "latex", booktabs = T, digits = 0,
      caption = paste0("Top 50 ", region_label, " Estimates, Range (EPest-low and EPest-high), 1-Yr, ", recent_year)) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r table-3yr, include=T}
kable(head(region_w$three %>% arrange(RANK_3YR) %>%
             select(RANK_3YR, COMPOUND, EPEST_LOW_KG_3YR_AVG, EPEST_HIGH_KG_3YR_AVG), n = 50),
      "latex", booktabs = T, digits = 0,
      caption = paste0("Top 50 ", region_label, " Estimates, Range (EPest-low and EPest-high), 3-Yr Avg, ",
                        paste(range(tail(window_years,3)), collapse="-"))) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r table-5yr, include=T}
kable(head(region_w$five %>% arrange(RANK_5YR) %>%
             select(RANK_5YR, COMPOUND, EPEST_LOW_KG_5YR_AVG, EPEST_HIGH_KG_5YR_AVG), n = 50),
      "latex", booktabs = T, digits = 0,
      caption = paste0("Top 50 ", region_label, " Estimates, Range (EPest-low and EPest-high), 5-Yr Avg, ",
                        paste(range(window_years), collapse="-"))) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```
