---
title: "Pacific Northwest Pesticide Class Use Trends, 1992-2018"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: bootstrap
    source_code: embed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_pkgs <- c("flexdashboard", "dplyr", "tidyr", "readr", "forcats", "stringr",
                    "ggplot2", "plotly", "DT", "crosstalk", "scales", "knitr", "kableExtra")
invisible(lapply(required_pkgs, function(pkg) {
  library(pkg, character.only = TRUE)
}))

options(scipen = 999)
```

```{r constants}
# ---- Report scope -----------------------------------------------------
# Washington, Oregon, Idaho -- the region this chapter summarizes. Order
# here fixes the order everywhere else (palette, legends, tables).
region_fips  <- c("53", "41", "16")
region_names <- c("53" = "Washington", "41" = "Oregon", "16" = "Idaho")
state_levels <- unname(region_names[region_fips])
region_label <- "Pacific Northwest (WA/OR/ID)"

crop_cols <- c("Corn", "Soybeans", "Wheat", "Cotton", "Vegetables_and_fruit",
                "Rice", "Orchards_and_grapes", "Alfalfa", "Pasture_and_hay",
                "Other_crops")

# ---- Palette (validated categorical order; see dataviz skill) ---------
cat_hex <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
              "#e87ba4", "#008300", "#4a3aa7", "#e34948")
muted_grey <- "#898781"

# TYPE is the broad, ~7-value grouping used for headline charts (fixed
# categorical color per type, "Other" folded to neutral grey). CLASS is
# the ~30-value chemical-family grouping (organophosphate, pyrethrin/pyrethroid,
# triazine, ...) used for filtering/drilldown, not as a simultaneous
# multi-series color encoding. STATE is a third, region-specific dimension
# (only 3 values, so it's safe to use as a chart color too -- see the State
# Comparison page).
type_levels <- c("Insecticide", "Herbicide", "Fungicide", "Fumigant",
                  "Plant Growth Regulator", "Biological", "Other")
type_colors <- setNames(c(cat_hex[1:6], muted_grey), type_levels)
state_colors <- setNames(cat_hex[seq_along(state_levels)], state_levels)

seq_blue <- c("#cde2fb", "#9ec5f4", "#5598e7", "#2a78d6", "#184f95")
```

```{r load-data}
raw <- read_delim(
  "../estimates/HighEstimate_AgPestUsebyCropGroup92to19.txt",
  delim = "\t",
  col_types = cols(
    State_FIPS_code = col_character(),
    State           = col_character(),
    Compound        = col_character(),
    Year            = col_integer(),
    Units           = col_character(),
    .default        = col_double()
  )
)

classification <- read_csv("../data/pesticide_classification.csv", col_types = cols())

# 2019 is excluded: this file's 2019 slice is a partial/preliminary release
# covering only 69 compounds nationwide (vs. ~400-410 in every other year),
# and that shortfall hits WA, OR, and ID alike. Including 2019 would draw a
# sharp artificial cliff at the end of every trend line. See the regional
# ranking chapter for the same exclusion.
pnw <- raw %>%
  filter(State_FIPS_code %in% region_fips, Year <= 2018) %>%
  inner_join(classification, by = c("Compound" = "COMPOUND")) %>%
  mutate(
    TYPE = factor(TYPE, levels = type_levels),
    STATE = factor(region_names[State_FIPS_code], levels = state_levels)
  )

pnw_long <- pnw %>%
  pivot_longer(cols = all_of(crop_cols), names_to = "CROP_GROUP", values_to = "EPEST_HIGH_KG") %>%
  mutate(
    EPEST_HIGH_KG = replace_na(EPEST_HIGH_KG, 0),
    CROP_GROUP = str_replace_all(CROP_GROUP, "_", " ")
  ) %>%
  filter(!is.na(Year)) %>%
  select(Year, STATE, Compound, CLASS, TYPE, CROP_GROUP, EPEST_HIGH_KG)

# Fixed, stable crop-group order (by total 1992-2018 mass, region-wide) so
# a given crop group keeps the same color everywhere in the dashboard.
# Crop groups with zero recorded use across the whole region and period
# are dropped rather than shown as empty series.
crop_group_totals <- pnw_long %>%
  group_by(CROP_GROUP) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  filter(TOTAL_KG > 0) %>%
  arrange(desc(TOTAL_KG))

crop_group_levels <- crop_group_totals$CROP_GROUP
crop_group_colors <- setNames(cat_hex[seq_along(crop_group_levels)], crop_group_levels)

pnw_long <- pnw_long %>%
  filter(CROP_GROUP %in% crop_group_levels) %>%
  mutate(CROP_GROUP = factor(CROP_GROUP, levels = crop_group_levels))

# Types with zero recorded use are dropped from TYPE-faceted charts the
# same way crop groups are.
type_totals <- pnw_long %>%
  group_by(TYPE) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  filter(TOTAL_KG > 0) %>%
  arrange(match(TYPE, type_levels))
active_types <- as.character(type_totals$TYPE)

year_range <- range(pnw_long$Year)
latest_year <- max(pnw_long$Year)
first_year <- min(pnw_long$Year)

# ---- Pre-aggregated summaries used throughout the dashboard -----------
annual_by_type <- pnw_long %>%
  group_by(Year, TYPE) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_by_type_crop <- pnw_long %>%
  filter(TYPE %in% active_types) %>%
  group_by(Year, TYPE, CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  mutate(TYPE = factor(TYPE, levels = active_types))

annual_by_compound <- pnw_long %>%
  group_by(Year, TYPE, CLASS, Compound) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_overall <- pnw_long %>%
  group_by(Year) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

# ---- State-comparison summaries (this chapter's reason to exist) ------
annual_by_state <- pnw_long %>%
  group_by(Year, STATE) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_by_state_type <- pnw_long %>%
  filter(TYPE %in% active_types) %>%
  group_by(Year, STATE, TYPE) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  mutate(TYPE = factor(TYPE, levels = active_types))
```

```{r overview-stats}
total_latest <- annual_overall %>% filter(Year == latest_year) %>% pull(EPEST_HIGH_KG)
total_first  <- annual_overall %>% filter(Year == first_year)  %>% pull(EPEST_HIGH_KG)
total_pct_change <- 100 * (total_latest - total_first) / total_first

top_type_latest <- annual_by_type %>%
  filter(Year == latest_year) %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)

top_crop_latest <- annual_by_type_crop %>%
  filter(Year == latest_year) %>%
  group_by(CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)

top_state_latest <- annual_by_state %>%
  filter(Year == latest_year) %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)
```

Overview
=======================================================================

Row {data-height=140}
-----------------------------------------------------------------------

### Total pesticide use, `r latest_year` {.value-box}

```{r}
valueBox(paste0(comma(round(total_latest)), " kg"),
         caption = paste0("All classified compounds, ", latest_year, " (EPest-high)"),
         icon = "fa-flask", color = cat_hex[1])
```

### Top type, `r latest_year` {.value-box}

```{r}
valueBox(paste0(as.character(top_type_latest$TYPE), ": ", comma(round(top_type_latest$EPEST_HIGH_KG)), " kg"),
         caption = paste0("Largest pesticide type by mass, ", latest_year),
         icon = "fa-layer-group", color = cat_hex[2])
```

### Change since `r first_year` {.value-box}

```{r}
valueBox(paste0(ifelse(total_pct_change >= 0, "+", ""), round(total_pct_change), "%"),
         caption = paste0("Total classified pesticide use, ", first_year, "-", latest_year),
         icon = ifelse(total_pct_change >= 0, "fa-arrow-up", "fa-arrow-down"),
         # Explicit hex, not the "warning" keyword: flexdashboard's default
         # theme pairs "warning" with white value-box text on a pale yellow
         # background, which is nearly unreadable.
         color = ifelse(total_pct_change >= 0, cat_hex[2], "success"))
```

### Top state, `r latest_year` {.value-box}

```{r}
valueBox(paste0(as.character(top_state_latest$STATE), ": ", comma(round(top_state_latest$EPEST_HIGH_KG)), " kg"),
         caption = paste0("Largest pesticide use by state, ", latest_year),
         icon = "fa-map", color = "primary")
```

### Top crop group, `r latest_year` {.value-box}

```{r}
valueBox(as.character(top_crop_latest$CROP_GROUP),
         caption = paste0("Largest combined use across all classes, ", latest_year),
         icon = "fa-seedling", color = "primary")
```

Row {data-height=430}
-----------------------------------------------------------------------

### Overall annual use — all classified pesticides combined, `r first_year`-`r latest_year`

```{r}
p <- plot_ly(annual_overall, x = ~Year, y = ~EPEST_HIGH_KG, type = "scatter", mode = "lines",
             fill = "tozeroy", fillcolor = "rgba(42,120,214,0.15)",
             line = list(color = cat_hex[1], width = 2),
             hovertemplate = "%{x}: %{y:,.0f} kg<extra></extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    hovermode = "x unified"
  )
p
```

Row {data-height=430}
-----------------------------------------------------------------------

### Annual use by pesticide type, `r first_year`-`r latest_year`

```{r}
plot_ly(annual_by_type, x = ~Year, y = ~EPEST_HIGH_KG, color = ~TYPE,
        colors = type_colors, type = "scatter", mode = "lines+markers",
        line = list(width = 2), marker = list(size = 5),
        hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    legend = list(title = list(text = "Type"), orientation = "h", y = -0.2),
    hovermode = "x unified"
  )
```

State Comparison
=======================================================================

Row {data-height=480}
-----------------------------------------------------------------------

### Annual use by state, `r first_year`-`r latest_year`

```{r}
plot_ly(annual_by_state, x = ~Year, y = ~EPEST_HIGH_KG, color = ~STATE,
        colors = state_colors, type = "scatter", mode = "lines+markers",
        line = list(width = 2), marker = list(size = 5),
        hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    legend = list(title = list(text = "State"), orientation = "h", y = -0.2),
    hovermode = "x unified"
  )
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (state x type view)

Tip: none of the mass totals below are shares of the regional total, and
the three states farm very different total acreages, so raw kg is not a
fair "which state uses more" comparison on its own (see the caveat on the
Methodology page). A faceted per-state-by-type static chart is on the
Static Plots page.

```{r}
datatable(
  annual_by_state_type,
  extensions = "Scroller",
  colnames = c("Year", "State", "Type", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Interactive Portal
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by crop group

Tip: lines are colored by crop group and dashed by type, **summed across
all three states** (see the State Comparison page to split by state
instead). Class-level filtering (30 chemical classes) is on the Compound
Explorer page instead — plotting crop-group trends split by all 30
classes at once would make the lines and legend unreadable, so this view
stops at type (7 categories).

```{r portal-crop-setup}
# Grain matches exactly what's plotted (Year x Type x Crop group): unlike
# a compound-level chart, a crop-group line has no compound identity to
# split on, so if the shared data still had a Class (or State) dimension
# underneath, multiple values' rows for the same year/crop-group would
# connect into one jagged, meaningless line. Aggregating away Class and
# State here (State Comparison and Compound Explorer cover those levels)
# keeps every plotted line a true 1-row-per-year series regardless of
# which Type boxes are checked.
type_crop_active <- pnw_long %>%
  filter(TYPE %in% active_types) %>%
  group_by(TYPE, CROP_GROUP) %>%
  mutate(PAIR_TOTAL = sum(EPEST_HIGH_KG)) %>%
  ungroup() %>%
  filter(PAIR_TOTAL > 0) %>%
  group_by(Year, TYPE, CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

crop_shared <- SharedData$new(type_crop_active, ~paste(Year, TYPE, CROP_GROUP), group = "crop_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("type_crop", "Type", crop_shared, ~TYPE, inline = TRUE),
      filter_select("crop_sel", "Crop group", crop_shared, ~CROP_GROUP, multiple = TRUE),
      filter_slider("year_crop", "Year", crop_shared, ~Year, step = 1, sep = "", width = "100%")
    ),
    plot_ly(crop_shared, x = ~Year, y = ~EPEST_HIGH_KG, color = ~CROP_GROUP,
            colors = crop_group_colors, linetype = ~TYPE,
            type = "scatter", mode = "lines+markers", height = 600,
            line = list(width = 2), marker = list(size = 5),
            hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year"),
        legend = list(title = list(text = "Crop group / type"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (crop group view)

```{r}
datatable(
  crop_shared,
  extensions = "Scroller",
  colnames = c("Year", "Type", "Crop group", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Compound Explorer
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by compound

Each line is one compound (color and hover label identify it), **summed
across all three states**. Narrow the Compound filter (or the Type/Class
filters) to a handful of active ingredients for the clearest view.

```{r portal-compound-setup}
compound_shared <- SharedData$new(annual_by_compound, ~paste(Year, TYPE, CLASS, Compound), group = "compound_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("type_cmp", "Type", compound_shared, ~TYPE, inline = TRUE),
      filter_select("class_cmp", "Class", compound_shared, ~CLASS, multiple = TRUE),
      filter_select("compound_sel", "Compound", compound_shared, ~Compound, multiple = TRUE),
      filter_slider("year_cmp", "Year", compound_shared, ~Year, step = 1, sep = "", width = "100%")
    ),
    plot_ly(compound_shared, x = ~Year, y = ~EPEST_HIGH_KG, color = ~Compound,
            text = ~paste0(TYPE, " / ", CLASS), height = 600,
            type = "scatter", mode = "lines+markers",
            line = list(width = 1.5), marker = list(size = 4),
            hovertemplate = "%{fullData.name} (%{text}), %{x}: %{y:,.0f} kg<extra></extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year"),
        legend = list(title = list(text = "Compound"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (compound view)

```{r}
datatable(
  compound_shared,
  extensions = "Scroller",
  colnames = c("Year", "Type", "Class", "Compound", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Static Plots
=======================================================================

Row {data-height=500}
-----------------------------------------------------------------------

### Overall annual use by type

```{r static-type-trend, fig.width=10, fig.height=5}
ggplot(annual_by_type, aes(x = Year, y = EPEST_HIGH_KG, color = TYPE)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = type_colors, name = "Type") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 2)) +
  labs(
    title = paste0(region_label, ": Annual pesticide use by type, ", first_year, "-", latest_year),
    subtitle = "USGS EPest-high estimates, summed across WA + OR + ID and all reported crop groups",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

### Annual use by state

```{r static-state-trend, fig.width=10, fig.height=5}
ggplot(annual_by_state, aes(x = Year, y = EPEST_HIGH_KG, color = STATE)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = state_colors, name = "State") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 2)) +
  labs(
    title = paste0(region_label, ": Annual pesticide use by state, ", first_year, "-", latest_year),
    subtitle = "USGS EPest-high estimates, all classified compounds and crop groups",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

Row {data-height=500}
-----------------------------------------------------------------------

### Top 20 compounds by cumulative use, `r first_year`-`r latest_year`

```{r static-top-compounds, fig.width=10, fig.height=5}
top_compounds <- annual_by_compound %>%
  group_by(TYPE, CLASS, Compound) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(TOTAL_KG)) %>%
  slice(1:20) %>%
  mutate(Compound = fct_reorder(Compound, TOTAL_KG))

ggplot(top_compounds, aes(x = Compound, y = TOTAL_KG, fill = TYPE)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = type_colors, name = "Type") +
  scale_y_continuous(labels = comma) +
  labs(
    title = paste0("Top 20 compounds region-wide by cumulative mass applied, ", first_year, "-", latest_year),
    x = NULL, y = "Cumulative mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

### Annual use by state, faceted by type

```{r static-state-by-type, fig.width=10, fig.height=5}
ggplot(annual_by_state_type, aes(x = Year, y = EPEST_HIGH_KG, color = STATE)) +
  geom_line(linewidth = 1) +
  facet_wrap(~TYPE, ncol = 2, scales = "free_y") +
  scale_color_manual(values = state_colors, name = "State") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6)) +
  labs(
    title = "Annual use by state, faceted by type",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Row {data-height=650}
-----------------------------------------------------------------------

### Annual use by crop group, faceted by type (region-wide)

```{r static-crop-area, fig.width=10, fig.height=7}
ggplot(annual_by_type_crop, aes(x = Year, y = EPEST_HIGH_KG, fill = CROP_GROUP)) +
  geom_area(position = "stack") +
  facet_wrap(~TYPE, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = crop_group_colors, name = "Crop group") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6)) +
  labs(
    title = "Annual use by crop group, faceted by type",
    subtitle = "Stacked EPest-high estimates, summed across WA + OR + ID; crop group order fixed by total 1992-2018 mass",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Row {data-height=650}
-----------------------------------------------------------------------

### Use intensity by year and crop group (region-wide)

```{r static-heatmap, fig.width=10, fig.height=7}
kg_breaks <- c(0, 10, 100, 1000, 10000, 100000, 1000000)
kg_breaks <- kg_breaks[kg_breaks <= max(annual_by_type_crop$EPEST_HIGH_KG)]

ggplot(annual_by_type_crop, aes(x = Year, y = CROP_GROUP, fill = log1p(EPEST_HIGH_KG))) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~TYPE, ncol = 2) +
  scale_fill_gradientn(colors = seq_blue, name = "kg",
                        breaks = log1p(kg_breaks), labels = comma(kg_breaks)) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6), expand = c(0, 0)) +
  labs(
    title = "Use intensity by year and crop group (log scale)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Methodology & Data
=======================================================================

Row
-----------------------------------------------------------------------

### About this dashboard

Data source: USGS Pesticide National Synthesis Project, *Estimated Annual
Agricultural Pesticide Use for Counties/States, by Crop Group* (1992-2019),
file `HighEstimate_AgPestUsebyCropGroup92to19.txt`, downloaded from
[ScienceBase item 6081ae7cd34e8564d6866222](https://www.sciencebase.gov/catalog/item/6081ae7cd34e8564d6866222).
**This dashboard uses 1992-2018 only** — the file's 2019 slice is a
partial/preliminary release covering just 69 compounds nationwide versus
roughly 400 in every other year, for all three states alike. With the
full classification in play, most compounds have no 2019 row at all, which
would draw a sharp artificial drop at the end of every trend line if
included. Re-run once a complete 2019 release is available.

This dashboard covers **`r region_label`** — Washington (FIPS 53), Oregon
(FIPS 41), and Idaho (FIPS 16) summed together, plus a dedicated State
Comparison page for the three states side by side. It's built from the
same `r nrow(classification)`-compound classification
(`data/pesticide_classification.csv`) as each state's own chapter.

**Comparing states by raw mass is not apples to apples.** Washington,
Oregon, and Idaho farm very different total acreages and crop mixes, so a
higher kg total for one state does not by itself mean heavier *use
intensity* — it may just mean more farmland. This chapter reports absolute
mass only; it does not normalize by acreage, and none of the charts here
should be read as "State X is dirtier than State Y" without that context.

**Three grouping columns.** `TYPE` is the broad functional category
(Insecticide, Herbicide, Fungicide, Fumigant, Plant Growth Regulator,
Biological, Other) used for the headline charts on the Overview and Static
Plots pages. `CLASS` is the specific chemical family (organophosphate,
pyrethrin/pyrethroid, triazine, DMI fungicide, ...) used for filtering on
the Interactive Portal and Compound Explorer pages — there are too many
classes to plot as simultaneous colored series, so they're a filter
dimension rather than a chart aesthetic. `STATE` (only 3 values) is safe
to use as a direct chart color and appears on the State Comparison and
Static Plots pages.

**EPest-high method.** Values are the *EPest-high* estimates described by
Baker and Stone (2015) and Thelin and Stone (2013): where a Crop Reporting
District (CRD) was surveyed but use of a given pesticide-by-crop combination
was not reported, EPest-high treats that as missing data and imputes a rate
from neighboring/regional CRDs, rather than assuming zero use (EPest-low).
This dataset provides EPest-high only — EPest-low is not available in this
crop-group product.

**Crop groups.** USGS aggregates crop-specific use into 10 groups: Corn,
Soybeans, Wheat, Cotton, Vegetables & fruit, Rice, Orchards & grapes, Alfalfa,
Pasture & hay, and Other crops. Groups with zero recorded region-wide use
across 1992-2018 for the classified compounds are omitted from the charts
above.

**Caveats (per USGS):**

* Reliability decreases with scale (e.g. "detailed interpretation of where
  and how much use occurs within a county is not appropriate"); summing
  three states together inherits this caveat at an even larger scale.
* Recent years use projected county/state crop acreage from the most recent
  Census of Agriculture at time of estimation and are revised as new Census
  data become available.
* A blank cell in the source file (no reported/estimated use for that
  compound-year-crop-group) is treated as zero in this dashboard.
* Classification into chemical classes is necessarily a simplification —
  a handful of compounds have real claims to more than one class (e.g.
  fenoxycarb is a carbamate insect growth regulator; propamocarb is a
  carbamic-acid-ester fungicide, not grouped with the insecticide/herbicide
  carbamates). Each compound is assigned to the single class judged most
  useful for a chemical-class trends view.

### Pesticide classification

`r nrow(classification)` compounds classified into `r n_distinct(classification$CLASS)`
chemical classes and `r n_distinct(classification$TYPE)` broad types.
Compounds reported per state: `r knitr::combine_words(paste0(state_levels, " (", sapply(state_levels, function(s) n_distinct(pnw$Compound[pnw$STATE == s])), ")"))`.

```{r classification-table}
classification %>%
  arrange(TYPE, CLASS, COMPOUND) %>%
  select(COMPOUND, CLASS, TYPE) %>%
  datatable(
    colnames = c("Compound", "Class", "Type"),
    options = list(pageLength = 15, dom = "ftip"),
    rownames = FALSE
  )
```
