---
title: "ID County-Level Pesticide Use, 2014-2018"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: bootstrap
    source_code: embed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_pkgs <- c("flexdashboard", "dplyr", "tidyr", "readr", "forcats", "stringr",
                    "ggplot2", "plotly", "DT", "crosstalk", "scales", "knitr", "kableExtra",
                    "maps", "sf")
invisible(lapply(required_pkgs, function(pkg) {
  library(pkg, character.only = TRUE)
}))

options(scipen = 999)
```

```{r constants}
# ---- Report scope -----------------------------------------------------
state_fips  <- "16"
state_name  <- "Idaho"
class_trends_file <- "id_pesticide_class_trends.html"

# Same 2014-2018 window as the statewide ranking chapter (not the full
# 1992-2018 sweep the trends dashboard uses): the county-level source is
# split one file per year, so reading all 27 years for a chapter that
# exists to complement the ranking chapter isn't worth the render time.
window_years <- 2014:2018
recent_year  <- max(window_years)

cat_hex <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
              "#e87ba4", "#008300", "#4a3aa7", "#e34948",
              "#8c564b", "#c25aab")
muted_grey <- "#898781"
functional_levels <- c("Herbicide/algicide", "Fungicide", "Insecticide",
                        "Plant Growth Regulator", "Acaricide", "Fumigant",
                        "Insect Growth Regulator", "Other Bactericides",
                        "Nematicides", "Insect Repellent", "Other")
functional_colors <- setNames(c(cat_hex[1:10], muted_grey), functional_levels)
seq_blue <- c("#cde2fb", "#9ec5f4", "#5598e7", "#2a78d6", "#184f95")
```

```{r load-data}
read_year <- function(yr) {
  read_delim(
    sprintf("../estimates/EPest.county.estimates.%d.txt", yr),
    delim = "\t",
    col_types = cols(
      COMPOUND = col_character(),
      YEAR = col_integer(),
      STATE_FIPS_CODE = col_character(),
      COUNTY_FIPS_CODE = col_character(),
      EPEST_LOW_KG = col_double(),
      EPEST_HIGH_KG = col_double()
    )
  ) %>%
    filter(STATE_FIPS_CODE == state_fips)
}

county_raw <- bind_rows(lapply(window_years, read_year))

classification <- read_csv("../data/pesticide_classification.csv", col_types = cols())

# ---- County name / geometry lookup (offline, from the "maps" package) -
# county.fips maps a 5-digit FIPS to a "state,county" name, but splits
# multi-part counties (islands, etc.) into several rows per FIPS with a
# ":subregion" suffix (e.g. Washington's San Juan county becomes
# "washington,san juan:lopez island", "washington,san juan:orcas island",
# etc.) -- the drawn polygons merge those back into one shape per county,
# so the ":..." suffix has to be stripped before joining to get back to
# one row per FIPS.
data(county.fips, envir = environment())
fips_lookup <- county.fips %>%
  mutate(FIPS = sprintf("%05d", fips), polyname = sub(":.*$", "", polyname)) %>%
  distinct(FIPS, polyname) %>%
  filter(startsWith(polyname, paste0(tolower(state_name), ","))) %>%
  mutate(COUNTY_NAME = tools::toTitleCase(sub("^[^,]+,", "", polyname)))

state_map <- st_as_sf(map("county", region = tolower(state_name), plot = FALSE, fill = TRUE)) %>%
  rename(polyname = ID)

county_long <- county_raw %>%
  inner_join(classification, by = c("COMPOUND" = "COMPOUND")) %>%
  mutate(
    FUNCTIONAL_CLASS = factor(FUNCTIONAL_CLASS, levels = functional_levels),
    FIPS = paste0(STATE_FIPS_CODE, COUNTY_FIPS_CODE)
  ) %>%
  left_join(fips_lookup, by = "FIPS") %>%
  select(YEAR, FIPS, COUNTY_NAME, polyname, CHEMICAL_CLASS, FUNCTIONAL_CLASS, COMPOUND, EPEST_HIGH_KG)

n_unmatched_fips <- county_long %>% filter(is.na(COUNTY_NAME)) %>% distinct(FIPS) %>% nrow()

# ---- Pre-aggregated summaries ------------------------------------------
annual_by_county <- county_long %>%
  filter(!is.na(COUNTY_NAME)) %>%
  group_by(YEAR, FIPS, COUNTY_NAME, polyname) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG, na.rm = TRUE), .groups = "drop")

annual_by_county_functional <- county_long %>%
  filter(!is.na(COUNTY_NAME)) %>%
  group_by(YEAR, FIPS, COUNTY_NAME, FUNCTIONAL_CLASS) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG, na.rm = TRUE), .groups = "drop")

by_compound_county <- county_long %>%
  filter(!is.na(COUNTY_NAME)) %>%
  group_by(YEAR, FIPS, COUNTY_NAME, FUNCTIONAL_CLASS, CHEMICAL_CLASS, COMPOUND) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG, na.rm = TRUE), .groups = "drop")

annual_state_total <- annual_by_county %>%
  group_by(YEAR) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

# Fixed, stable county order (by total mass across the window) so a given
# county keeps the same rank position and the "top N" charts are stable.
county_totals_window <- annual_by_county %>%
  group_by(FIPS, COUNTY_NAME) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(TOTAL_KG))

# 1/3/5-yr county ranking, same independent-per-window approach as the
# statewide pesticide ranking chapter (a county's presence in one window's
# ranking doesn't depend on its presence in the others).
rank_window <- function(yrs) {
  annual_by_county %>%
    filter(YEAR %in% yrs) %>%
    group_by(FIPS, COUNTY_NAME) %>%
    summarize(EPEST_HIGH_KG = mean(EPEST_HIGH_KG), .groups = "drop") %>%
    arrange(desc(EPEST_HIGH_KG)) %>%
    mutate(RANK = row_number())
}

rank_1yr <- rank_window(recent_year) %>% rename(EPEST_HIGH_KG_1YR = EPEST_HIGH_KG, RANK_1YR = RANK)
rank_3yr <- rank_window(tail(window_years, 3)) %>% rename(EPEST_HIGH_KG_3YR_AVG = EPEST_HIGH_KG, RANK_3YR = RANK)
rank_5yr <- rank_window(window_years) %>% rename(EPEST_HIGH_KG_5YR_AVG = EPEST_HIGH_KG, RANK_5YR = RANK)

county_ranked <- rank_1yr %>%
  full_join(rank_3yr %>% select(FIPS, EPEST_HIGH_KG_3YR_AVG, RANK_3YR), by = "FIPS") %>%
  full_join(rank_5yr %>% select(FIPS, EPEST_HIGH_KG_5YR_AVG, RANK_5YR), by = "FIPS") %>%
  arrange(RANK_1YR)

n_counties_total <- nrow(fips_lookup)
n_counties_active <- n_distinct(annual_by_county$FIPS)
```

```{r overview-stats}
top_county_latest <- annual_by_county %>%
  filter(YEAR == recent_year) %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)

state_total_latest <- annual_state_total %>% filter(YEAR == recent_year) %>% pull(EPEST_HIGH_KG)
top_county_share <- 100 * top_county_latest$EPEST_HIGH_KG / state_total_latest

top5_share <- 100 * sum(county_totals_window$TOTAL_KG[1:5]) / sum(county_totals_window$TOTAL_KG)
```

Overview
=======================================================================

Row {data-height=90}
-----------------------------------------------------------------------

### About this page {.no-title}

<div style="padding: 10px 16px; background:#f9f9f7; border-left: 4px solid #898781; font-size: 0.9em;">
<strong>Read this before the map.</strong> USGS's own documentation notes
these are <strong>county-level estimates whose primary intended use is to
be aggregated into state-level totals</strong> — not necessarily precise,
independently reliable figures for any one county. Neighboring counties
can differ in these numbers partly because of how use was modeled and
allocated across Crop Reporting Districts, not only because of real
differences in what growers applied. Use this page to see what's
<em>driving</em> `r state_name`'s statewide totals (Chapters 2-3), not as a
precise county-by-county scoreboard. See the Methodology page for the
full caveat.
</div>

Row {data-height=140}
-----------------------------------------------------------------------

### Top county, `r recent_year` {.value-box}

```{r}
valueBox(paste0(top_county_latest$COUNTY_NAME, ": ", comma(round(top_county_latest$EPEST_HIGH_KG)), " kg"),
         caption = paste0("Largest classified pesticide use by county, ", recent_year),
         icon = "fa-map-marker-alt", color = cat_hex[1])
```

### Top county's share {.value-box}

```{r}
valueBox(paste0(round(top_county_share), "%"),
         caption = paste0("Share of statewide ", recent_year, " total in the top county alone"),
         icon = "fa-percent", color = cat_hex[2])
```

### Top 5 counties' share {.value-box}

```{r}
valueBox(paste0(round(top5_share), "%"),
         caption = paste0("Share of ", first(window_years), "-", recent_year, " total mass in the top 5 counties"),
         icon = "fa-layer-group", color = "primary")
```

### Counties with recorded use {.value-box}

```{r}
valueBox(paste0(n_counties_active, " / ", n_counties_total),
         caption = paste0(state_name, " counties reporting classified pesticide use, ", first(window_years), "-", recent_year),
         icon = "fa-map", color = muted_grey)
```

Row {data-height=600}
-----------------------------------------------------------------------

### `r recent_year` county totals (EPest-high, log scale)

```{r static-map-overview, fig.width=8, fig.height=6.5}
map_data_latest <- state_map %>%
  left_join(annual_by_county %>% filter(YEAR == recent_year), by = "polyname")

ggplot(map_data_latest) +
  geom_sf(aes(fill = EPEST_HIGH_KG), color = "white", linewidth = 0.2) +
  scale_fill_gradientn(colors = seq_blue, name = "kg", trans = "log1p",
                        breaks = scales::breaks_log(n = 5),
                        labels = comma, na.value = "grey90") +
  labs(title = paste0(state_name, " county totals, ", recent_year),
       subtitle = "All classified compounds, EPest-high; grey = no recorded use") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")
```

### Top 10 counties, `r recent_year`

```{r}
top10 <- annual_by_county %>%
  filter(YEAR == recent_year) %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1:10) %>%
  mutate(COUNTY_NAME = fct_reorder(COUNTY_NAME, EPEST_HIGH_KG))

plot_ly(top10, x = ~EPEST_HIGH_KG, y = ~COUNTY_NAME, type = "bar", orientation = "h",
        marker = list(color = cat_hex[1]),
        hovertemplate = "%{y}: %{x:,.0f} kg<extra></extra>") %>%
  layout(
    xaxis = list(title = "EPest-high estimate (kg)", tickformat = ","),
    yaxis = list(title = "")
  )
```

County Ranking
=======================================================================

Row {data-height=600}
-----------------------------------------------------------------------

### Full county ranking

```{r}
display_rank <- county_ranked %>%
  mutate(across(starts_with("EPEST"), ~round(.x))) %>%
  select(RANK_1YR, COUNTY_NAME, EPEST_HIGH_KG_1YR, RANK_3YR, EPEST_HIGH_KG_3YR_AVG, RANK_5YR, EPEST_HIGH_KG_5YR_AVG)

datatable(
  display_rank,
  colnames = c(paste0("Rank (", recent_year, ")"), "County", paste0("EPest-high ", recent_year, " (kg)"),
               "Rank (3yr)", "EPest-high 3yr avg (kg)", "Rank (5yr)", "EPest-high 5yr avg (kg)"),
  options = list(pageLength = 15, dom = "ftip", order = list(list(0, "asc"))),
  rownames = FALSE
) %>%
  formatRound(c("EPEST_HIGH_KG_1YR", "EPEST_HIGH_KG_3YR_AVG", "EPEST_HIGH_KG_5YR_AVG"), 0)
```

Row {data-height=500}
-----------------------------------------------------------------------

### Top 15 counties by cumulative use, `r first(window_years)`-`r recent_year`

```{r static-top-counties, fig.width=10, fig.height=5}
top15_cum <- county_totals_window %>%
  slice(1:15) %>%
  mutate(COUNTY_NAME = fct_reorder(COUNTY_NAME, TOTAL_KG))

ggplot(top15_cum, aes(x = COUNTY_NAME, y = TOTAL_KG)) +
  geom_col(fill = cat_hex[1]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = paste0("Top 15 counties by cumulative mass applied, ", first(window_years), "-", recent_year),
    x = NULL, y = "Cumulative EPest-high mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

### Trend for the top 6 counties

```{r}
top6_fips <- county_totals_window$FIPS[1:6]

trend_top6 <- annual_by_county %>%
  filter(FIPS %in% top6_fips) %>%
  mutate(COUNTY_NAME = factor(COUNTY_NAME, levels = county_totals_window$COUNTY_NAME[1:6]))

plot_ly(trend_top6, x = ~YEAR, y = ~EPEST_HIGH_KG, color = ~COUNTY_NAME, colors = cat_hex[1:6],
        type = "scatter", mode = "lines+markers", line = list(width = 2), marker = list(size = 5),
        hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 1),
    legend = list(title = list(text = "County"))
  )
```

County Explorer
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by county

Each line is one county, dashed by functional class (color identifies the
county, hover label gives the exact functional class). With
`r n_counties_active` counties reporting use across 11 functional classes,
narrow the County and/or Functional class filters to a handful for the
clearest view — lines are always split by county x functional class under
the hood, so no combination of filters connects mismatched points into a
jagged line.

```{r portal-county-setup}
# Grain matches exactly what's plotted (Year x FIPS x Functional class): a
# line is only ever one county's one functional class, so leaving either
# filter unfiltered can't connect mismatched points the way it would if
# functional class were aggregated away.
county_shared <- SharedData$new(annual_by_county_functional, ~paste(YEAR, FIPS, FUNCTIONAL_CLASS), group = "county_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("func_county", "Functional class", county_shared, ~FUNCTIONAL_CLASS, inline = TRUE),
      filter_select("county_sel", "County", county_shared, ~COUNTY_NAME, multiple = TRUE),
      filter_slider("year_county", "Year", county_shared, ~YEAR, step = 1, sep = "", width = "100%")
    ),
    plot_ly(county_shared, x = ~YEAR, y = ~EPEST_HIGH_KG, color = ~COUNTY_NAME, linetype = ~FUNCTIONAL_CLASS,
            height = 600, type = "scatter", mode = "lines+markers",
            line = list(width = 1.5), marker = list(size = 4),
            hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year", dtick = 1),
        legend = list(title = list(text = "County / functional class"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (county view)

```{r}
datatable(
  county_shared,
  colnames = c("Year", "FIPS", "County", "Functional class", "EPest-high (kg)"),
  options = list(pageLength = 10, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Compound Explorer
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by compound, within a county

**Pick a county first.** Each line is one compound, but the underlying data
has a row per county — with no county selected (or several selected at
once), a single compound's line connects every county's separate value
year to year, which is *not* a statewide total and will look like a
meaningless jagged zig-zag. Narrowing County to one (or a couple, for
direct comparison) keeps every line a true single-place series. For an
already-aggregated statewide-combined view by compound, see the [Pesticide
Class Trends](`r class_trends_file`) chapter's own Compound
Explorer instead.

```{r portal-compound-setup}
compound_shared <- SharedData$new(by_compound_county, ~paste(YEAR, FIPS, FUNCTIONAL_CLASS, CHEMICAL_CLASS, COMPOUND), group = "compound_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_select("county_cmp", "County", compound_shared, ~COUNTY_NAME, multiple = TRUE),
      filter_checkbox("func_cmp", "Functional class", compound_shared, ~FUNCTIONAL_CLASS, inline = TRUE),
      filter_select("class_cmp", "Chemical class", compound_shared, ~CHEMICAL_CLASS, multiple = TRUE),
      filter_select("compound_sel", "Compound", compound_shared, ~COMPOUND, multiple = TRUE),
      filter_slider("year_cmp", "Year", compound_shared, ~YEAR, step = 1, sep = "", width = "100%")
    ),
    plot_ly(compound_shared, x = ~YEAR, y = ~EPEST_HIGH_KG, color = ~COMPOUND, height = 650,
            text = ~paste0(COUNTY_NAME, ": ", FUNCTIONAL_CLASS, " / ", CHEMICAL_CLASS),
            type = "scatter", mode = "lines+markers",
            line = list(width = 1.5), marker = list(size = 4),
            hovertemplate = "%{fullData.name} (%{text}), %{x}: %{y:,.0f} kg<extra></extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year", dtick = 1),
        legend = list(title = list(text = "Compound"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (compound x county view)

```{r}
datatable(
  compound_shared,
  colnames = c("Year", "FIPS", "County", "Functional class", "Chemical class", "Compound", "EPest-high (kg)"),
  options = list(pageLength = 10, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Static Plots
=======================================================================

Row {data-height=600}
-----------------------------------------------------------------------

### `r first(window_years)` county totals (EPest-high, log scale)

```{r static-map-first, fig.width=8, fig.height=6.5}
map_data_first <- state_map %>%
  left_join(annual_by_county %>% filter(YEAR == first(window_years)), by = "polyname")

ggplot(map_data_first) +
  geom_sf(aes(fill = EPEST_HIGH_KG), color = "white", linewidth = 0.2) +
  scale_fill_gradientn(colors = seq_blue, name = "kg", trans = "log1p",
                        breaks = scales::breaks_log(n = 5),
                        labels = comma, na.value = "grey90") +
  labs(title = paste0(state_name, " county totals, ", first(window_years)),
       subtitle = "All classified compounds, EPest-high; grey = no recorded use") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")
```

### Change in county totals, `r first(window_years)`-`r recent_year`

```{r static-map-change, fig.width=8, fig.height=6.5}
change_data <- annual_by_county %>%
  filter(YEAR %in% c(first(window_years), recent_year)) %>%
  select(FIPS, COUNTY_NAME, polyname, YEAR, EPEST_HIGH_KG) %>%
  pivot_wider(names_from = YEAR, values_from = EPEST_HIGH_KG, names_prefix = "Y") %>%
  mutate(PCT_CHANGE = 100 * (.data[[paste0("Y", recent_year)]] - .data[[paste0("Y", first(window_years))]]) /
           .data[[paste0("Y", first(window_years))]])

map_data_change <- state_map %>% left_join(change_data, by = "polyname")

lim <- max(abs(map_data_change$PCT_CHANGE), na.rm = TRUE)
ggplot(map_data_change) +
  geom_sf(aes(fill = PCT_CHANGE), color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2a78d6", mid = "#f0efec", high = "#e34948", midpoint = 0,
                        limits = c(-lim, lim), name = "% change", labels = comma, na.value = "grey90") +
  labs(title = paste0("Change in county totals, ", first(window_years), " to ", recent_year),
       subtitle = "Percent change, EPest-high; grey = no use in one or both years") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")
```

Row {data-height=500}
-----------------------------------------------------------------------

### Annual use by county, top 8 counties, by functional class

```{r static-county-type, fig.width=10, fig.height=5}
top8_fips <- county_totals_window$FIPS[1:8]

annual_by_county_functional %>%
  filter(FIPS %in% top8_fips) %>%
  mutate(COUNTY_NAME = factor(COUNTY_NAME, levels = county_totals_window$COUNTY_NAME[1:8])) %>%
  ggplot(aes(x = YEAR, y = EPEST_HIGH_KG, fill = FUNCTIONAL_CLASS)) +
  geom_col(position = "stack") +
  facet_wrap(~COUNTY_NAME, ncol = 4) +
  scale_fill_manual(values = functional_colors, name = "Functional class") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = window_years) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
  labs(
    title = "Annual use by functional class, top 8 counties",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 8),
    axis.text.x = element_text(angle = 90, size = 7)
  )
```

Methodology & Data
=======================================================================

Row
-----------------------------------------------------------------------

### About this dashboard

Data source: USGS Pesticide National Synthesis Project, *Estimated Annual
Agricultural Pesticide Use for Counties* (`estimates/EPest.county.estimates.<year>.txt`),
the same county-level files summed to a state total in the
`r state_name` statewide top-pesticide ranking chapter — this dashboard
shows the county-level detail *behind* those state totals rather than
re-deriving them.

**Read the caveat, seriously.** USGS's own documentation for this product
says: *"we report county-level estimates despite their intended use to
generate state-level estimates."* In plain terms: these numbers were
designed and validated as inputs to a state-level total, not as
precise, independently reliable figures for any one county. Two specific
consequences for this dashboard:

* Differences between neighboring counties can reflect how the
  proprietary Crop Reporting District (CRD) survey rates were allocated
  across the counties within a CRD, rather than a real difference in what
  growers applied.
* "Detailed interpretation of where and how much use occurs within a
  county is not appropriate" (the same caution given in every other
  chapter in this repo) applies doubly once county is itself the unit of
  analysis.

Use this chapter to see *what's driving* `r state_name`'s statewide totals
(which counties, which pesticide functional classes) — not as a precise
county-by-county scoreboard, and not to draw conclusions about specific
farms or fields.

This dashboard covers **`r state_name` (FIPS `r state_fips`)**,
`r first(window_years)`-`r recent_year` (same window as the statewide
ranking chapter, for direct comparability), filtered to the
`r nrow(classification)` compounds classified in
`data/pesticide_classification.csv`. `r n_counties_active` of
`r n_counties_total` `r state_name` counties reported at least one
classified compound in this window.

**County geometry.** County boundaries and names come from R's `maps`
package (`maps::county.fips`, offline, no county-level crop-group data is
available from USGS at this granularity — the crop-group breakdown used
in the pesticide class trends chapter is only published at the state
level). `r ifelse(n_unmatched_fips == 0, "Every county-FIPS code in the source data matched a boundary.", paste(n_unmatched_fips, "county-FIPS code(s) in the source data did not match a known boundary and are excluded from the maps."))`

**EPest-high method.** Values are the *EPest-high* estimates described by
Baker and Stone (2015) and Thelin and Stone (2013) — see the statewide
ranking chapter for the full description. This chapter does not use
EPest-low.

**Caveats (per USGS), in addition to the one above:**

* Recent years use projected county crop acreage from the most recent
  Census of Agriculture at time of estimation and are revised as new
  Census data become available.
* A blank cell in the source file (no reported/estimated use for that
  compound-year-county) is treated as zero in this dashboard.
* 2019 is not included in this chapter's window for the same reason it's
  excluded from the other chapters: that year's file is a
  partial/preliminary release covering far fewer compounds than 2014-2018.

### Pesticide classification

`r nrow(classification)` compounds classified into `r n_distinct(classification$CHEMICAL_CLASS)`
chemical classes and `r n_distinct(classification$FUNCTIONAL_CLASS)` functional
classes; see the [Pesticide Class Trends](`r class_trends_file`) chapter for the full table.

```{r classification-table}
classification %>%
  arrange(FUNCTIONAL_CLASS, CHEMICAL_CLASS, COMPOUND) %>%
  select(COMPOUND, CHEMICAL_CLASS, FUNCTIONAL_CLASS) %>%
  datatable(
    colnames = c("Compound", "Chemical class", "Functional class"),
    options = list(pageLength = 10, dom = "ftip"),
    rownames = FALSE
  )
```
