---
title: "ID Pesticide Class Use Trends, 1992-2018"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: bootstrap
    source_code: embed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_pkgs <- c("flexdashboard", "dplyr", "tidyr", "readr", "forcats", "stringr",
                    "ggplot2", "plotly", "DT", "crosstalk", "scales", "knitr", "kableExtra")
invisible(lapply(required_pkgs, function(pkg) {
  library(pkg, character.only = TRUE)
}))

options(scipen = 999)
```

```{r constants}
# ---- Report scope -----------------------------------------------------
state_fips <- "16"
state_name <- "Idaho"

crop_cols <- c("Corn", "Soybeans", "Wheat", "Cotton", "Vegetables_and_fruit",
                "Rice", "Orchards_and_grapes", "Alfalfa", "Pasture_and_hay",
                "Other_crops")

# ---- Palette (validated categorical order; see dataviz skill) ---------
cat_hex <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
              "#e87ba4", "#008300", "#4a3aa7", "#e34948")
muted_grey <- "#898781"

# TYPE is the broad, ~7-value grouping used for headline charts (fixed
# categorical color per type, "Other" folded to neutral grey). CLASS is
# the ~30-value chemical-family grouping (organophosphate, pyrethroid,
# triazine, ...) used for filtering/drilldown, not as a simultaneous
# multi-series color encoding.
type_levels <- c("Insecticide", "Herbicide", "Fungicide", "Fumigant",
                  "Plant Growth Regulator", "Biological", "Other")
type_colors <- setNames(c(cat_hex[1:6], muted_grey), type_levels)

seq_blue <- c("#cde2fb", "#9ec5f4", "#5598e7", "#2a78d6", "#184f95")
```

```{r load-data}
raw <- read_delim(
  "../estimates/HighEstimate_AgPestUsebyCropGroup92to19.txt",
  delim = "\t",
  col_types = cols(
    State_FIPS_code = col_character(),
    State           = col_character(),
    Compound        = col_character(),
    Year            = col_integer(),
    Units           = col_character(),
    .default        = col_double()
  )
)

classification <- read_csv("../data/pesticide_classification.csv", col_types = cols())

# 2019 is excluded: this file's 2019 slice is a partial/preliminary release
# covering only 69 compounds nationwide (vs. ~400-410 in every other year).
# With the full pesticide-classification taxonomy in play (rather than the
# ~36-compound organophosphate/carbamate subset used previously, most of
# which happen to be in that partial release), including 2019 would draw a
# sharp artificial cliff at the end of every trend line. See the
# statewide-ranking chapter for the same exclusion.
wa <- raw %>%
  filter(State_FIPS_code == state_fips, Year <= 2018) %>%
  inner_join(classification, by = c("Compound" = "COMPOUND")) %>%
  mutate(TYPE = factor(TYPE, levels = type_levels))

wa_long <- wa %>%
  pivot_longer(cols = all_of(crop_cols), names_to = "CROP_GROUP", values_to = "EPEST_HIGH_KG") %>%
  mutate(
    EPEST_HIGH_KG = replace_na(EPEST_HIGH_KG, 0),
    CROP_GROUP = str_replace_all(CROP_GROUP, "_", " ")
  ) %>%
  filter(!is.na(Year)) %>%
  select(Year, Compound, CLASS, TYPE, CROP_GROUP, EPEST_HIGH_KG)

# Fixed, stable crop-group order (by total 1992-2018 mass) so a given crop
# group keeps the same color everywhere in the dashboard. Crop groups with
# zero recorded use across the whole period are dropped rather than shown
# as empty series.
crop_group_totals <- wa_long %>%
  group_by(CROP_GROUP) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  filter(TOTAL_KG > 0) %>%
  arrange(desc(TOTAL_KG))

crop_group_levels <- crop_group_totals$CROP_GROUP
crop_group_colors <- setNames(cat_hex[seq_along(crop_group_levels)], crop_group_levels)

wa_long <- wa_long %>%
  filter(CROP_GROUP %in% crop_group_levels) %>%
  mutate(CROP_GROUP = factor(CROP_GROUP, levels = crop_group_levels))

# Types with zero recorded use are dropped from TYPE-faceted charts the
# same way crop groups are.
type_totals <- wa_long %>%
  group_by(TYPE) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  filter(TOTAL_KG > 0) %>%
  arrange(match(TYPE, type_levels))
active_types <- as.character(type_totals$TYPE)

year_range <- range(wa_long$Year)
latest_year <- max(wa_long$Year)
first_year <- min(wa_long$Year)

# ---- Pre-aggregated summaries used throughout the dashboard -----------
annual_by_type <- wa_long %>%
  group_by(Year, TYPE) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_by_type_crop <- wa_long %>%
  filter(TYPE %in% active_types) %>%
  group_by(Year, TYPE, CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  mutate(TYPE = factor(TYPE, levels = active_types))

annual_by_compound <- wa_long %>%
  group_by(Year, TYPE, CLASS, Compound) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_overall <- wa_long %>%
  group_by(Year) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")
```

```{r overview-stats}
total_latest <- annual_overall %>% filter(Year == latest_year) %>% pull(EPEST_HIGH_KG)
total_first  <- annual_overall %>% filter(Year == first_year)  %>% pull(EPEST_HIGH_KG)
total_pct_change <- 100 * (total_latest - total_first) / total_first

top_type_latest <- annual_by_type %>%
  filter(Year == latest_year) %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)

top_crop_latest <- annual_by_type_crop %>%
  filter(Year == latest_year) %>%
  group_by(CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)
```

Overview
=======================================================================

Row {data-height=140}
-----------------------------------------------------------------------

### Total pesticide use, `r latest_year` {.value-box}

```{r}
valueBox(paste0(comma(round(total_latest)), " kg"),
         caption = paste0("All classified compounds, ", latest_year, " (EPest-high)"),
         icon = "fa-flask", color = cat_hex[1])
```

### Top type, `r latest_year` {.value-box}

```{r}
valueBox(paste0(as.character(top_type_latest$TYPE), ": ", comma(round(top_type_latest$EPEST_HIGH_KG)), " kg"),
         caption = paste0("Largest pesticide type by mass, ", latest_year),
         icon = "fa-layer-group", color = cat_hex[2])
```

### Change since `r first_year` {.value-box}

```{r}
valueBox(paste0(ifelse(total_pct_change >= 0, "+", ""), round(total_pct_change), "%"),
         caption = paste0("Total classified pesticide use, ", first_year, "-", latest_year),
         icon = ifelse(total_pct_change >= 0, "fa-arrow-up", "fa-arrow-down"),
         # Explicit hex, not the "warning" keyword: flexdashboard's default
         # theme pairs "warning" with white value-box text on a pale yellow
         # background, which is nearly unreadable (see dataviz skill's
         # status-palette contrast notes). cat_hex[2] keeps white-text
         # contrast on an increase; "success" already renders fine.
         color = ifelse(total_pct_change >= 0, cat_hex[2], "success"))
```

### Top crop group, `r latest_year` {.value-box}

```{r}
valueBox(as.character(top_crop_latest$CROP_GROUP),
         caption = paste0("Largest combined use across all classes, ", latest_year),
         icon = "fa-seedling", color = "primary")
```

Row {data-height=430}
-----------------------------------------------------------------------

### Overall annual use — all classified pesticides combined, `r first_year`-`r latest_year`

```{r}
p <- plot_ly(annual_overall, x = ~Year, y = ~EPEST_HIGH_KG, type = "scatter", mode = "lines",
             fill = "tozeroy", fillcolor = "rgba(42,120,214,0.15)",
             line = list(color = cat_hex[1], width = 2),
             hovertemplate = "%{x}: %{y:,.0f} kg<extra></extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    hovermode = "x unified"
  )
p
```

Row {data-height=430}
-----------------------------------------------------------------------

### Annual use by pesticide type, `r first_year`-`r latest_year`

```{r}
plot_ly(annual_by_type, x = ~Year, y = ~EPEST_HIGH_KG, color = ~TYPE,
        colors = type_colors, type = "scatter", mode = "lines+markers",
        line = list(width = 2), marker = list(size = 5),
        hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    legend = list(title = list(text = "Type"), orientation = "h", y = -0.2),
    hovermode = "x unified"
  )
```

Interactive Portal
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by crop group

Tip: lines are colored by crop group and dashed by type. Class-level
filtering (30 chemical classes) is on the Compound Explorer page instead —
plotting crop-group trends split by all 30 classes at once would make the
lines and legend unreadable, so this view stops at type (7 categories).

```{r portal-crop-setup}
# Grain matches exactly what's plotted (Year x Type x Crop group): unlike
# a compound-level chart, a crop-group line has no compound identity to
# split on, so if the shared data still had a Class dimension underneath,
# multiple classes' values for the same year/crop-group would connect into
# one jagged, meaningless line. Aggregating away Class here (Compound
# Explorer covers that level) keeps every plotted line a true 1-row-per-year
# series regardless of which Type boxes are checked.
type_crop_active <- wa_long %>%
  filter(TYPE %in% active_types) %>%
  group_by(TYPE, CROP_GROUP) %>%
  mutate(PAIR_TOTAL = sum(EPEST_HIGH_KG)) %>%
  ungroup() %>%
  filter(PAIR_TOTAL > 0) %>%
  group_by(Year, TYPE, CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

crop_shared <- SharedData$new(type_crop_active, ~paste(Year, TYPE, CROP_GROUP), group = "crop_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("type_crop", "Type", crop_shared, ~TYPE, inline = TRUE),
      filter_select("crop_sel", "Crop group", crop_shared, ~CROP_GROUP, multiple = TRUE),
      filter_slider("year_crop", "Year", crop_shared, ~Year, step = 1, sep = "", width = "100%")
    ),
    plot_ly(crop_shared, x = ~Year, y = ~EPEST_HIGH_KG, color = ~CROP_GROUP,
            colors = crop_group_colors, linetype = ~TYPE,
            type = "scatter", mode = "lines+markers", height = 600,
            line = list(width = 2), marker = list(size = 5),
            hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year"),
        legend = list(title = list(text = "Crop group / type"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (crop group view)

```{r}
datatable(
  crop_shared,
  extensions = "Scroller",
  colnames = c("Year", "Type", "Crop group", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Compound Explorer
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by compound

Each line is one compound (color and hover label identify it). Narrow the
Compound filter (or the Type/Class filters) to a handful of active
ingredients for the clearest view.

```{r portal-compound-setup}
compound_shared <- SharedData$new(annual_by_compound, ~paste(Year, TYPE, CLASS, Compound), group = "compound_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("type_cmp", "Type", compound_shared, ~TYPE, inline = TRUE),
      filter_select("class_cmp", "Class", compound_shared, ~CLASS, multiple = TRUE),
      filter_select("compound_sel", "Compound", compound_shared, ~Compound, multiple = TRUE),
      filter_slider("year_cmp", "Year", compound_shared, ~Year, step = 1, sep = "", width = "100%")
    ),
    plot_ly(compound_shared, x = ~Year, y = ~EPEST_HIGH_KG, color = ~Compound,
            text = ~paste0(TYPE, " / ", CLASS), height = 600,
            type = "scatter", mode = "lines+markers",
            line = list(width = 1.5), marker = list(size = 4),
            hovertemplate = "%{fullData.name} (%{text}), %{x}: %{y:,.0f} kg<extra></extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year"),
        legend = list(title = list(text = "Compound"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (compound view)

```{r}
datatable(
  compound_shared,
  extensions = "Scroller",
  colnames = c("Year", "Type", "Class", "Compound", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Static Plots
=======================================================================

Row {data-height=500}
-----------------------------------------------------------------------

### Overall annual use by type

```{r static-type-trend, fig.width=10, fig.height=5}
ggplot(annual_by_type, aes(x = Year, y = EPEST_HIGH_KG, color = TYPE)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = type_colors, name = "Type") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 2)) +
  labs(
    title = paste0(state_name, ": Annual pesticide use by type, ", first_year, "-", latest_year),
    subtitle = "USGS EPest-high estimates, summed across all reported crop groups",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

### Top 20 compounds by cumulative use, `r first_year`-`r latest_year`

```{r static-top-compounds, fig.width=10, fig.height=5}
top_compounds <- annual_by_compound %>%
  group_by(TYPE, CLASS, Compound) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(TOTAL_KG)) %>%
  slice(1:20) %>%
  mutate(Compound = fct_reorder(Compound, TOTAL_KG))

ggplot(top_compounds, aes(x = Compound, y = TOTAL_KG, fill = TYPE)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = type_colors, name = "Type") +
  scale_y_continuous(labels = comma) +
  labs(
    title = paste0("Top 20 compounds by cumulative mass applied, ", first_year, "-", latest_year),
    x = NULL, y = "Cumulative mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

Row {data-height=650}
-----------------------------------------------------------------------

### Annual use by crop group

```{r static-crop-area, fig.width=10, fig.height=7}
ggplot(annual_by_type_crop, aes(x = Year, y = EPEST_HIGH_KG, fill = CROP_GROUP)) +
  geom_area(position = "stack") +
  facet_wrap(~TYPE, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = crop_group_colors, name = "Crop group") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6)) +
  labs(
    title = "Annual use by crop group, faceted by type",
    subtitle = "Stacked EPest-high estimates; crop group order fixed by total 1992-2018 mass",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Row {data-height=650}
-----------------------------------------------------------------------

### Use intensity by year and crop group

```{r static-heatmap, fig.width=10, fig.height=7}
kg_breaks <- c(0, 10, 100, 1000, 10000, 100000, 1000000)
kg_breaks <- kg_breaks[kg_breaks <= max(annual_by_type_crop$EPEST_HIGH_KG)]

ggplot(annual_by_type_crop, aes(x = Year, y = CROP_GROUP, fill = log1p(EPEST_HIGH_KG))) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~TYPE, ncol = 2) +
  scale_fill_gradientn(colors = seq_blue, name = "kg",
                        breaks = log1p(kg_breaks), labels = comma(kg_breaks)) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6), expand = c(0, 0)) +
  labs(
    title = "Use intensity by year and crop group (log scale)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Methodology & Data
=======================================================================

Row
-----------------------------------------------------------------------

### About this dashboard

Data source: USGS Pesticide National Synthesis Project, *Estimated Annual
Agricultural Pesticide Use for Counties/States, by Crop Group* (1992-2019),
file `HighEstimate_AgPestUsebyCropGroup92to19.txt`, downloaded from
[ScienceBase item 6081ae7cd34e8564d6866222](https://www.sciencebase.gov/catalog/item/6081ae7cd34e8564d6866222).
**This dashboard uses 1992-2018 only** — the file's 2019 slice is a
partial/preliminary release covering just 69 compounds nationwide (48 for
`r state_name`) versus roughly 400 (220+) in every other year. With the
full classification in play, most compounds have no 2019 row at all, which
would draw a sharp artificial drop at the end of every trend line if
included. Re-run once a complete 2019 release is available.

This dashboard covers **`r state_name` (FIPS `r state_fips`)** only,
filtered to the `r nrow(classification)` compounds classified in
`data/pesticide_classification.csv` (see the table below). Every compound
reported for `r state_name` in the source file is classified; none are
dropped as "unclassifiable" — the classification's own `Other/Unclassified`
class is reserved for the small number of miscellaneous compounds (a few
adjuvants, a molluscicide, a nitrogen stabilizer) that don't belong to any
recognized pesticide class.

**Two grouping columns.** `TYPE` is the broad functional category
(Insecticide, Herbicide, Fungicide, Fumigant, Plant Growth Regulator,
Biological, Other) used for the headline charts on the Overview and Static
Plots pages. `CLASS` is the specific chemical family (organophosphate,
pyrethroid, triazine, DMI fungicide, ...) used for filtering on the
Interactive Portal and Compound Explorer pages — there are too many
classes to plot as simultaneous colored series, so they're a filter
dimension rather than a chart aesthetic.

**EPest-high method.** Values are the *EPest-high* estimates described by
Baker and Stone (2015) and Thelin and Stone (2013): where a Crop Reporting
District (CRD) was surveyed but use of a given pesticide-by-crop combination
was not reported, EPest-high treats that as missing data and imputes a rate
from neighboring/regional CRDs, rather than assuming zero use (EPest-low).
This dataset provides EPest-high only — EPest-low is not available in this
crop-group product.

**Crop groups.** USGS aggregates crop-specific use into 10 groups: Corn,
Soybeans, Wheat, Cotton, Vegetables & fruit, Rice, Orchards & grapes, Alfalfa,
Pasture & hay, and Other crops. Groups with zero recorded `r state_name` use
across 1992-2018 for the classified compounds are omitted from the charts
above.

**Caveats (per USGS):**

* Recent years use projected county/state crop acreage from the most recent
  Census of Agriculture at time of estimation and are revised as new Census
  data become available.
* Estimates are not appropriate for interpreting where/how much use occurs
  within a crop group at fine spatial detail; they summarize survey-based
  application rates aggregated to the state and crop-group level.
* A blank cell in the source file (no reported/estimated use for that
  compound-year-crop-group) is treated as zero in this dashboard.
* Classification into chemical classes is necessarily a simplification —
  a handful of compounds have real claims to more than one class (e.g.
  fenoxycarb is a carbamate insect growth regulator; propamocarb is a
  carbamic-acid-ester fungicide, not grouped with the insecticide/herbicide
  carbamates). Each compound is assigned to the single class judged most
  useful for a chemical-class trends view.

### Pesticide classification

`r nrow(classification)` compounds classified into `r n_distinct(classification$CLASS)`
chemical classes and `r n_distinct(classification$TYPE)` broad types;
`r n_distinct(wa$Compound)` of those were reported for `r state_name`.

```{r classification-table}
classification %>%
  arrange(TYPE, CLASS, COMPOUND) %>%
  select(COMPOUND, CLASS, TYPE) %>%
  datatable(
    colnames = c("Compound", "Class", "Type"),
    options = list(pageLength = 15, dom = "ftip"),
    rownames = FALSE
  )
```
