---
title: "ID Pesticide Class Use Trends, 1992-2018"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: bootstrap
    source_code: embed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_pkgs <- c("flexdashboard", "dplyr", "tidyr", "readr", "forcats", "stringr",
                    "ggplot2", "plotly", "DT", "crosstalk", "scales", "knitr", "kableExtra")
invisible(lapply(required_pkgs, function(pkg) {
  library(pkg, character.only = TRUE)
}))

options(scipen = 999)
```

```{r constants}
# ---- Report scope -----------------------------------------------------
state_fips <- "16"
state_name <- "Idaho"

crop_cols <- c("Corn", "Soybeans", "Wheat", "Cotton", "Vegetables_and_fruit",
                "Rice", "Orchards_and_grapes", "Alfalfa", "Pasture_and_hay",
                "Other_crops")

# ---- Palette (validated categorical order; see dataviz skill) ---------
cat_hex <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
              "#e87ba4", "#008300", "#4a3aa7", "#e34948",
              "#8c564b", "#c25aab")
muted_grey <- "#898781"

# FUNCTIONAL_CLASS is the broad, use-based grouping (what a compound is used
# for: Insecticide, Herbicide/algicide, Fungicide, Fumigant, ...) used for
# headline charts (fixed categorical color per class, "Other" folded to
# neutral grey). CHEMICAL_CLASS is the structural/chemical-family grouping
# (Organophosphorous compound, Pyrethrins/Pyrethroids, Triazines, ...) used
# for filtering/drilldown, not as a simultaneous multi-series color
# encoding -- most compounds' chemistry falls outside this taxonomy's named
# families and lands in "Other", so it isn't tractable as a chart color
# either. See the Methodology page for how each was assigned.
functional_levels <- c("Herbicide/algicide", "Fungicide", "Insecticide",
                        "Plant Growth Regulator", "Acaricide", "Fumigant",
                        "Insect Growth Regulator", "Other Bactericides",
                        "Nematicides", "Insect Repellent", "Other")
functional_colors <- setNames(c(cat_hex[1:10], muted_grey), functional_levels)

seq_blue <- c("#cde2fb", "#9ec5f4", "#5598e7", "#2a78d6", "#184f95")
```

```{r load-data}
raw <- read_delim(
  "../estimates/HighEstimate_AgPestUsebyCropGroup92to19.txt",
  delim = "\t",
  col_types = cols(
    State_FIPS_code = col_character(),
    State           = col_character(),
    Compound        = col_character(),
    Year            = col_integer(),
    Units           = col_character(),
    .default        = col_double()
  )
)

classification <- read_csv("../data/pesticide_classification.csv", col_types = cols())

# 2019 is excluded: this file's 2019 slice is a partial/preliminary release
# covering only 69 compounds nationwide (vs. ~400-410 in every other year).
# With the full pesticide-classification taxonomy in play (rather than the
# ~36-compound organophosphate/carbamate subset used previously, most of
# which happen to be in that partial release), including 2019 would draw a
# sharp artificial cliff at the end of every trend line. See the
# statewide-ranking chapter for the same exclusion.
wa <- raw %>%
  filter(State_FIPS_code == state_fips, Year <= 2018) %>%
  inner_join(classification, by = c("Compound" = "COMPOUND")) %>%
  mutate(FUNCTIONAL_CLASS = factor(FUNCTIONAL_CLASS, levels = functional_levels))

wa_long <- wa %>%
  pivot_longer(cols = all_of(crop_cols), names_to = "CROP_GROUP", values_to = "EPEST_HIGH_KG") %>%
  mutate(
    EPEST_HIGH_KG = replace_na(EPEST_HIGH_KG, 0),
    CROP_GROUP = str_replace_all(CROP_GROUP, "_", " ")
  ) %>%
  filter(!is.na(Year)) %>%
  select(Year, Compound, CHEMICAL_CLASS, FUNCTIONAL_CLASS, CROP_GROUP, EPEST_HIGH_KG)

# Fixed, stable crop-group order (by total 1992-2018 mass) so a given crop
# group keeps the same color everywhere in the dashboard. Crop groups with
# zero recorded use across the whole period are dropped rather than shown
# as empty series.
crop_group_totals <- wa_long %>%
  group_by(CROP_GROUP) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  filter(TOTAL_KG > 0) %>%
  arrange(desc(TOTAL_KG))

crop_group_levels <- crop_group_totals$CROP_GROUP
crop_group_colors <- setNames(cat_hex[seq_along(crop_group_levels)], crop_group_levels)

wa_long <- wa_long %>%
  filter(CROP_GROUP %in% crop_group_levels) %>%
  mutate(CROP_GROUP = factor(CROP_GROUP, levels = crop_group_levels))

# Functional classes with zero recorded use are dropped from
# FUNCTIONAL_CLASS-faceted charts the same way crop groups are.
functional_totals <- wa_long %>%
  group_by(FUNCTIONAL_CLASS) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  filter(TOTAL_KG > 0) %>%
  arrange(match(FUNCTIONAL_CLASS, functional_levels))
active_functional_classes <- as.character(functional_totals$FUNCTIONAL_CLASS)

year_range <- range(wa_long$Year)
latest_year <- max(wa_long$Year)
first_year <- min(wa_long$Year)

# ---- Pre-aggregated summaries used throughout the dashboard -----------
annual_by_functional <- wa_long %>%
  group_by(Year, FUNCTIONAL_CLASS) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_by_functional_crop <- wa_long %>%
  filter(FUNCTIONAL_CLASS %in% active_functional_classes) %>%
  group_by(Year, FUNCTIONAL_CLASS, CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  mutate(FUNCTIONAL_CLASS = factor(FUNCTIONAL_CLASS, levels = active_functional_classes))

annual_by_compound <- wa_long %>%
  group_by(Year, FUNCTIONAL_CLASS, CHEMICAL_CLASS, Compound) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

annual_overall <- wa_long %>%
  group_by(Year) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")
```

```{r overview-stats}
total_latest <- annual_overall %>% filter(Year == latest_year) %>% pull(EPEST_HIGH_KG)
total_first  <- annual_overall %>% filter(Year == first_year)  %>% pull(EPEST_HIGH_KG)
total_pct_change <- 100 * (total_latest - total_first) / total_first

top_functional_latest <- annual_by_functional %>%
  filter(Year == latest_year) %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)

top_crop_latest <- annual_by_functional_crop %>%
  filter(Year == latest_year) %>%
  group_by(CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(EPEST_HIGH_KG)) %>%
  slice(1)
```

Overview
=======================================================================

Row {data-height=140}
-----------------------------------------------------------------------

### Total pesticide use, `r latest_year` {.value-box}

```{r}
valueBox(paste0(comma(round(total_latest)), " kg"),
         caption = paste0("All classified compounds, ", latest_year, " (EPest-high)"),
         icon = "fa-flask", color = cat_hex[1])
```

### Top functional class, `r latest_year` {.value-box}

```{r}
valueBox(paste0(as.character(top_functional_latest$FUNCTIONAL_CLASS), ": ", comma(round(top_functional_latest$EPEST_HIGH_KG)), " kg"),
         caption = paste0("Largest pesticide functional class by mass, ", latest_year),
         icon = "fa-layer-group", color = cat_hex[2])
```

### Change since `r first_year` {.value-box}

```{r}
valueBox(paste0(ifelse(total_pct_change >= 0, "+", ""), round(total_pct_change), "%"),
         caption = paste0("Total classified pesticide use, ", first_year, "-", latest_year),
         icon = ifelse(total_pct_change >= 0, "fa-arrow-up", "fa-arrow-down"),
         # Explicit hex, not the "warning" keyword: flexdashboard's default
         # theme pairs "warning" with white value-box text on a pale yellow
         # background, which is nearly unreadable (see dataviz skill's
         # status-palette contrast notes). cat_hex[2] keeps white-text
         # contrast on an increase; "success" already renders fine.
         color = ifelse(total_pct_change >= 0, cat_hex[2], "success"))
```

### Top crop group, `r latest_year` {.value-box}

```{r}
valueBox(as.character(top_crop_latest$CROP_GROUP),
         caption = paste0("Largest combined use across all classes, ", latest_year),
         icon = "fa-seedling", color = "primary")
```

Row {data-height=430}
-----------------------------------------------------------------------

### Overall annual use — all classified pesticides combined, `r first_year`-`r latest_year`

```{r}
p <- plot_ly(annual_overall, x = ~Year, y = ~EPEST_HIGH_KG, type = "scatter", mode = "lines",
             fill = "tozeroy", fillcolor = "rgba(42,120,214,0.15)",
             line = list(color = cat_hex[1], width = 2),
             hovertemplate = "%{x}: %{y:,.0f} kg<extra></extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    hovermode = "x unified"
  )
p
```

Row {data-height=430}
-----------------------------------------------------------------------

### Annual use by functional class, `r first_year`-`r latest_year`

```{r}
plot_ly(annual_by_functional, x = ~Year, y = ~EPEST_HIGH_KG, color = ~FUNCTIONAL_CLASS,
        colors = functional_colors, type = "scatter", mode = "lines+markers",
        line = list(width = 2), marker = list(size = 5),
        hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
  layout(
    yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
    xaxis = list(title = "Year", dtick = 2),
    legend = list(title = list(text = "Functional class"), orientation = "h", y = -0.2),
    hovermode = "x unified"
  )
```

Interactive Portal
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by crop group

Tip: lines are colored by crop group and dashed by functional class.
Chemical-class filtering (13 named families, most compounds falling
outside them into "Other") is on the Compound Explorer page instead —
plotting crop-group trends split by chemical class at once would make the
lines and legend unreadable, so this view stops at functional class (11
categories).

```{r portal-crop-setup}
# Grain matches exactly what's plotted (Year x Functional class x Crop
# group): unlike a compound-level chart, a crop-group line has no compound
# identity to split on, so if the shared data still had a chemical-class
# dimension underneath, multiple classes' values for the same
# year/crop-group would connect into one jagged, meaningless line.
# Aggregating away chemical class here (Compound Explorer covers that
# level) keeps every plotted line a true 1-row-per-year series regardless
# of which functional-class boxes are checked.
functional_crop_active <- wa_long %>%
  filter(FUNCTIONAL_CLASS %in% active_functional_classes) %>%
  group_by(FUNCTIONAL_CLASS, CROP_GROUP) %>%
  mutate(PAIR_TOTAL = sum(EPEST_HIGH_KG)) %>%
  ungroup() %>%
  filter(PAIR_TOTAL > 0) %>%
  group_by(Year, FUNCTIONAL_CLASS, CROP_GROUP) %>%
  summarize(EPEST_HIGH_KG = sum(EPEST_HIGH_KG), .groups = "drop")

crop_shared <- SharedData$new(functional_crop_active, ~paste(Year, FUNCTIONAL_CLASS, CROP_GROUP), group = "crop_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("func_crop", "Functional class", crop_shared, ~FUNCTIONAL_CLASS, inline = TRUE),
      filter_select("crop_sel", "Crop group", crop_shared, ~CROP_GROUP, multiple = TRUE),
      filter_slider("year_crop", "Year", crop_shared, ~Year, step = 1, sep = "", width = "100%")
    ),
    plot_ly(crop_shared, x = ~Year, y = ~EPEST_HIGH_KG, color = ~CROP_GROUP,
            colors = crop_group_colors, linetype = ~FUNCTIONAL_CLASS,
            type = "scatter", mode = "lines+markers", height = 600,
            line = list(width = 2), marker = list(size = 5),
            hovertemplate = "%{x}: %{y:,.0f} kg<extra>%{fullData.name}</extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year"),
        legend = list(title = list(text = "Crop group / functional class"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (crop group view)

```{r}
datatable(
  crop_shared,
  extensions = "Scroller",
  colnames = c("Year", "Functional class", "Crop group", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Compound Explorer
=======================================================================

Row {data-height=700}
-----------------------------------------------------------------------

### Explore by compound

Each line is one compound (color and hover label identify it). Narrow the
Compound filter (or the Functional class/Chemical class filters) to a
handful of active ingredients for the clearest view.

```{r portal-compound-setup}
compound_shared <- SharedData$new(annual_by_compound, ~paste(Year, FUNCTIONAL_CLASS, CHEMICAL_CLASS, Compound), group = "compound_portal")
```

```{r}
htmltools::div(
  style = "padding: 8px 8px 8px 22px;",
  bscols(
    widths = c(3, 9),
    list(
      filter_checkbox("func_cmp", "Functional class", compound_shared, ~FUNCTIONAL_CLASS, inline = TRUE),
      filter_select("class_cmp", "Chemical class", compound_shared, ~CHEMICAL_CLASS, multiple = TRUE),
      filter_select("compound_sel", "Compound", compound_shared, ~Compound, multiple = TRUE),
      filter_slider("year_cmp", "Year", compound_shared, ~Year, step = 1, sep = "", width = "100%")
    ),
    plot_ly(compound_shared, x = ~Year, y = ~EPEST_HIGH_KG, color = ~Compound,
            text = ~paste0(FUNCTIONAL_CLASS, " / ", CHEMICAL_CLASS), height = 600,
            type = "scatter", mode = "lines+markers",
            line = list(width = 1.5), marker = list(size = 4),
            hovertemplate = "%{fullData.name} (%{text}), %{x}: %{y:,.0f} kg<extra></extra>") %>%
      layout(
        yaxis = list(title = "EPest-high estimate (kg)", rangemode = "tozero", tickformat = ","),
        xaxis = list(title = "Year"),
        legend = list(title = list(text = "Compound"), font = list(size = 9))
      )
  )
)
```

Row {data-height=400}
-----------------------------------------------------------------------

### Underlying data (compound view)

```{r}
datatable(
  compound_shared,
  extensions = "Scroller",
  colnames = c("Year", "Functional class", "Chemical class", "Compound", "EPest-high (kg)"),
  options = list(deferRender = TRUE, scrollY = 260, scroller = TRUE, dom = "ftip"),
  rownames = FALSE
) %>%
  formatRound("EPEST_HIGH_KG", 0)
```

Static Plots
=======================================================================

Row {data-height=500}
-----------------------------------------------------------------------

### Overall annual use by functional class

```{r static-type-trend, fig.width=10, fig.height=5}
ggplot(annual_by_functional, aes(x = Year, y = EPEST_HIGH_KG, color = FUNCTIONAL_CLASS)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = functional_colors, name = "Functional class") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 2)) +
  labs(
    title = paste0(state_name, ": Annual pesticide use by functional class, ", first_year, "-", latest_year),
    subtitle = "USGS EPest-high estimates, summed across all reported crop groups",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

### Top 20 compounds by cumulative use, `r first_year`-`r latest_year`

```{r static-top-compounds, fig.width=10, fig.height=5}
top_compounds <- annual_by_compound %>%
  group_by(FUNCTIONAL_CLASS, CHEMICAL_CLASS, Compound) %>%
  summarize(TOTAL_KG = sum(EPEST_HIGH_KG), .groups = "drop") %>%
  arrange(desc(TOTAL_KG)) %>%
  slice(1:20) %>%
  mutate(Compound = fct_reorder(Compound, TOTAL_KG))

ggplot(top_compounds, aes(x = Compound, y = TOTAL_KG, fill = FUNCTIONAL_CLASS)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = functional_colors, name = "Functional class") +
  scale_y_continuous(labels = comma) +
  labs(
    title = paste0("Top 20 compounds by cumulative mass applied, ", first_year, "-", latest_year),
    x = NULL, y = "Cumulative mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

Row {data-height=650}
-----------------------------------------------------------------------

### Annual use by crop group

```{r static-crop-area, fig.width=10, fig.height=7}
ggplot(annual_by_functional_crop, aes(x = Year, y = EPEST_HIGH_KG, fill = CROP_GROUP)) +
  geom_area(position = "stack") +
  facet_wrap(~FUNCTIONAL_CLASS, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = crop_group_colors, name = "Crop group") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6)) +
  labs(
    title = "Annual use by crop group, faceted by functional class",
    subtitle = "Stacked EPest-high estimates; crop group order fixed by total 1992-2018 mass",
    x = NULL, y = "Mass applied (kg)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Row {data-height=650}
-----------------------------------------------------------------------

### Use intensity by year and crop group

```{r static-heatmap, fig.width=10, fig.height=7}
kg_breaks <- c(0, 10, 100, 1000, 10000, 100000, 1000000)
kg_breaks <- kg_breaks[kg_breaks <= max(annual_by_functional_crop$EPEST_HIGH_KG)]

ggplot(annual_by_functional_crop, aes(x = Year, y = CROP_GROUP, fill = log1p(EPEST_HIGH_KG))) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~FUNCTIONAL_CLASS, ncol = 2) +
  scale_fill_gradientn(colors = seq_blue, name = "kg",
                        breaks = log1p(kg_breaks), labels = comma(kg_breaks)) +
  scale_x_continuous(breaks = seq(first_year, latest_year, by = 6), expand = c(0, 0)) +
  labs(
    title = "Use intensity by year and crop group (log scale)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

Methodology & Data
=======================================================================

Row
-----------------------------------------------------------------------

### About this dashboard

Data source: USGS Pesticide National Synthesis Project, *Estimated Annual
Agricultural Pesticide Use for Counties/States, by Crop Group* (1992-2019),
file `HighEstimate_AgPestUsebyCropGroup92to19.txt`, downloaded from
[ScienceBase item 6081ae7cd34e8564d6866222](https://www.sciencebase.gov/catalog/item/6081ae7cd34e8564d6866222).
**This dashboard uses 1992-2018 only** — the file's 2019 slice is a
partial/preliminary release covering just 69 compounds nationwide (48 for
`r state_name`) versus roughly 400 (220+) in every other year. With the
full classification in play, most compounds have no 2019 row at all, which
would draw a sharp artificial drop at the end of every trend line if
included. Re-run once a complete 2019 release is available.

This dashboard covers **`r state_name` (FIPS `r state_fips`)** only,
filtered to the `r nrow(classification)` compounds classified in
`data/pesticide_classification.csv` (see the table below). Every compound
reported for `r state_name` in the source file is classified; none are
dropped as "unclassifiable" — the classification's own `Other` bucket is
reserved for compounds outside the named families below (a few adjuvants,
a molluscicide, a nitrogen stabilizer, and — for Chemical Class
specifically — most modern chemistries not covered by the requested
taxonomy; see below).

**Two grouping columns, verified compound-by-compound.** USGS's own
metadata for these data releases (Identification/Entity-Attribute
Information sections) documents only the file's structural columns
(compound name, year, FIPS codes, crop-group amounts) — it does not
publish a compound-level chemical/functional classification crosswalk.
Both columns below were therefore built and checked one compound at a time
against pesticide chemistry/mode-of-action references, not derived from a
USGS-supplied table:

* **`FUNCTIONAL_CLASS`** — what the compound is *used for*: Insecticide,
  Herbicide/algicide, Fungicide, Fumigant, Rodenticide, Disinfectant/Broad
  Spectrum for Water Sanitation, Other Bactericides, Insect Repellent,
  Antifouling agent, Nematicides, Acaricides, Plant Growth Regulator
  (an addition beyond the base list, since ~19 compounds in this data are
  distinctly plant growth regulators rather than crop-protection products),
  or Other. Used for the headline charts on the Overview and Static Plots
  pages (only categories with recorded `r state_name` use are plotted;
  Rodenticide, Disinfectant, and Antifouling have zero compounds in this
  agricultural-use dataset).
* **`CHEMICAL_CLASS`** — the compound's structural/chemical family:
  Organochlorine compound, Organophosphorous compound, N-methyl carbamates
  (AChE inhibitor), Pyrethrins/Pyrethroids, Dipyridyl compound,
  Chlorophenoxy compound, Triazines, Thiocarbamates, Organo-metallic
  compound, Inorganic compounds, Microbial, Dithiocarbamates, or Other.
  This is a fairly classical taxonomy that predates several chemistries
  now in wide use (sulfonylurea and other ALS-inhibiting herbicides, DMI
  and SDHI and strobilurin fungicides, chloroacetanilide and dinitroaniline
  herbicides, neonicotinoid and diamide insecticides, and more) — those
  compounds correctly land in "Other" rather than being forced into an
  ill-fitting named family. `r nrow(classification %>% filter(CHEMICAL_CLASS == "Other"))`
  of `r nrow(classification)` compounds (`r round(100*nrow(classification %>% filter(CHEMICAL_CLASS == "Other"))/nrow(classification))`%)
  fall into "Other" for this reason. Used for filtering on the Interactive
  Portal and Compound Explorer pages — with most compounds in a single
  "Other" bucket and the rest split across 12 named families, it isn't a
  useful simultaneous chart-color dimension either.

Splitting these into two columns (rather than one column doing both jobs,
as in this dashboard's first version) directly fixes a real distortion: a
handful of fumigants (metam, metam potassium, dichloropropene,
chloropicrin) used to share a single "Fumigant" bucket that behaved like a
chemical class, when fumigants are actually chemically diverse (a
dithiocarbamate, in metam's case) and are applied at far higher per-acre
rates than herbicides or insecticides. That made raw-mass comparisons
across the old classes misleading — Fumigant could dwarf Herbicide in
total kg even though it covers only a handful of compounds. Functional
Class still correctly shows fumigants' true (large) total mass; Chemical
Class now distributes that same mass into the compounds' real chemistry
instead of an artificial catch-all.

**EPest-high method.** Values are the *EPest-high* estimates described by
Baker and Stone (2015) and Thelin and Stone (2013): where a Crop Reporting
District (CRD) was surveyed but use of a given pesticide-by-crop combination
was not reported, EPest-high treats that as missing data and imputes a rate
from neighboring/regional CRDs, rather than assuming zero use (EPest-low).
This dataset provides EPest-high only — EPest-low is not available in this
crop-group product.

**Crop groups.** USGS aggregates crop-specific use into 10 groups: Corn,
Soybeans, Wheat, Cotton, Vegetables & fruit, Rice, Orchards & grapes, Alfalfa,
Pasture & hay, and Other crops. Groups with zero recorded `r state_name` use
across 1992-2018 for the classified compounds are omitted from the charts
above.

**Caveats (per USGS):**

* Recent years use projected county/state crop acreage from the most recent
  Census of Agriculture at time of estimation and are revised as new Census
  data become available.
* Estimates are not appropriate for interpreting where/how much use occurs
  within a crop group at fine spatial detail; they summarize survey-based
  application rates aggregated to the state and crop-group level.
* A blank cell in the source file (no reported/estimated use for that
  compound-year-crop-group) is treated as zero in this dashboard.
* Classification is necessarily a simplification for compounds with
  legitimate dual uses — e.g. several compounds registered as both
  acaricides and general insecticides (abamectin, spiromesifen,
  spirotetramat) are assigned to whichever use is most commonly the
  compound's primary registered purpose; fenoxycarb is chemically a
  carbamate ester but mechanistically an insect growth regulator (not an
  AChE inhibitor), so it's classified by mechanism rather than by the
  "carbamate" name.

### Pesticide classification

`r nrow(classification)` compounds classified into `r n_distinct(classification$CHEMICAL_CLASS)`
chemical classes and `r n_distinct(classification$FUNCTIONAL_CLASS)` functional
classes; `r n_distinct(wa$Compound)` of those were reported for `r state_name`.

```{r classification-table}
classification %>%
  arrange(FUNCTIONAL_CLASS, CHEMICAL_CLASS, COMPOUND) %>%
  select(COMPOUND, CHEMICAL_CLASS, FUNCTIONAL_CLASS) %>%
  datatable(
    colnames = c("Compound", "Chemical class", "Functional class"),
    options = list(pageLength = 15, dom = "ftip"),
    rownames = FALSE
  )
```
